Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-10-03 Manuscript/manuscriptPS.tex
Update text

Two different definitions for the salt concentrations have been used when
electrometer concept is applied to study ion binding affinity.
The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. The used definition has only a marginal effect
to the results is simulations with realistic ion binding affinity.
\todo{Maybe we should add a figure in the SI.}

@@

Evaluation of counterion binding affinity to charged lipid bilayers
using the changes of headgroup order parameters is difficult because
reference state without ions does not exist, in contrast to previous
studies with zwitterionic membranes and coions \cite{catte16},
see also section \ref{mixtureTOadditionalCIs} in the supplementary information.
However, the counterion binding affinity is
reflected to the increase of POPC headgroup order parameters due to the
increasing amount of negatively charged POPS lipids in the bilayer.
The experimentally measured increase in the order parameters~\cite{scherer87}
is reproduced only in MacRog simulations, while other simulations
predict no change or even decrease in the order parameters (Fig. \ref{HGorderparametersPCvsPS}).
The results can be explained using the electrometer concept \cite{seelig87,scherer89}.
In experiments and MacRog simulations, the increasing amount of negative charge
in the membrane tilts the PC headgroup more parallel to the membrane plane
leading to the larger order parameters. In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged
lipids and the order parameters decrease with the increasing amount of
positive counterions of PS lipid.
In CHARMM36 simulations, the counterion binding neutralizes the negative charge
of lipids at the interface and the headgroup order parameters are not changed
with increasing amount of PS. Therefore, the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
response of POPC headgroup to the increasing amount of negatively charged lipids,
indicating that these simulations have realistic surface charge density in membranes
containing PS lipids.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

The headgroup order parameters of PS lipids
shift closer to zero when bilayer is diluted with PC lipids in 
experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.

Dilution of PS lipid bilayers with PC lipids reduces the propensity
of PS headgroup-multivalent ion complexes and is proposed to make PS
headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
Therefore, the intermolecular interactions at the headgroup region seems
to be important for the physical properties of mixed lipid bilayers.
These interactions can be indirectly monitored by measuring the
headgroup order parameters from PS:PC mixtures with different molar
ratios.

Oppositely to experiments, the headgroup order parameter of POPS shift away
from zero in CHARM36 simulations when bilayer is diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
In lipid14/17 simulations, the POPS order parameter shift closer to zero when
bilayer is diluted with POPC, but the numerical values of order parameters
are too far from experiments to enable interpretation of the experimental data.
Therefore, we conlcude that the force field development is necessary before
MD simulations can be used to interpret the interactions between PC and PS headgroups.
%The $\beta$-carbon order parameter
%increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
%to the experimental data.
%The smaller $\alpha$-carbon order parameter increase in both
%simulation models with increasing amount of PS, while it is almost unchanged in experiments.
%The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
%with increasing amount of PS, both model exhibiting a poor agreement with experiments.

>>>>

Two different definitions for the salt concentrations have been used when
electrometer concept is applied to study ion binding affinity.
The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. The used definition has only a marginal effect
to the results is simulations with realistic ion binding affinity
(section \ref{concentrationDEFsection} in the supplementary information).

@@

Evaluation of counterion binding affinity to charged lipid bilayers
using the changes of headgroup order parameters is difficult because
reference state without ions does not exist, in contrast to previous
studies with zwitterionic membranes and coions \cite{catte16},
see also section \ref{mixtureTOadditionalCIs} in the supplementary information.
However, the counterion binding affinity is
reflected to the increase of POPC headgroup order parameters due to the
increasing amount of negatively charged POPS lipids in the bilayer.
The experimentally measured increase in the order parameters~\cite{scherer87}
is reproduced only in MacRog simulations, while other simulations
predict no change or even decrease in the order parameters (Fig. \ref{HGorderparametersPCvsPS}).
The results can be explained using the electrometer concept \cite{seelig87,scherer89}.
In experiments and MacRog simulations, the increasing amount of negative charge
in the membrane tilts the PC headgroup more parallel to the membrane plane
leading to the larger order parameters. In Berger simulations,
the strong sodium binding surpasses the effect of negatively charged
lipids and the order parameters decrease with the increasing amount of
positive counterions of PS lipid.
In CHARMM36 simulations, the counterion binding neutralizes the negative charge
of lipids at the interface and the headgroup order parameters are not changed
with increasing amount of PS. Therefore, the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) reproduces the correct
response of POPC headgroup to the increasing amount of negatively charged lipids,
indicating that these simulations have realistic surface charge density in membranes
containing PS lipids.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure*}

Dilution of PS lipid bilayers with PC lipids reduces the propensity
of PS headgroup-multivalent ion complexes and is proposed to make PS
headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
This is related to the shift of the PS headgroup order parameters 
closer to zero when bilayer is diluted with PC
lipids (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90}.
%which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.
This shift is not seen in CHARM36 and Gromos-CKP simulations when bilayer is
diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
In lipid14/17 simulations, the headgroup order parameters of POPS shift closer
to zero when the bilayer is diluted with POPC, but the numerical values of
order parameters are too far from experiments, having also different signs,
for the proper interpretation of the experimental data.
Therefore, the imporvements in force fields are necessary
for accurate studies of interactions between PC and PS headgroups
with MD simulations.
%The $\beta$-carbon order parameter
%increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
%to the experimental data.
%The smaller $\alpha$-carbon order parameter increase in both
%simulation models with increasing amount of PS, while it is almost unchanged in experiments.
%The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
%with increasing amount of PS, both model exhibiting a poor agreement with experiments.

%Therefore, the intermolecular interactions at the headgroup region seems
%to be important for the physical properties of mixed lipid bilayers.
%These interactions can be indirectly monitored by measuring the
%headgroup order parameters from PS:PC mixtures with different molar
%ratios.

### 2018-10-03 Manuscript/manuscriptPSsuppl.tex
Update text

\pagebreak
\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFfigure}

>>>>

\pagebreak
\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
